---
title: "Group_Ass_2"
author: "Group_3"
date: "2024-10-30"
output: html_document
---

```{r,warning=FALSE}
library(fitdistrplus)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readxl)
```

```{r,warning=FALSE}
data <- read_excel("Patient_Data_2.xlsx")
data
```
```{r}
data <- data %>% 
  mutate(Date_Started=as.Date(Date_Started),Date_Ended=as.Date(Date_Ended))
data
```

# Over Viewing the Duration Column, Changing Numeric, and Using Absolute value for the delay days:
```{r}
data <- data %>% mutate(Duration=Date_Ended - Date_Started)
data$Duration<-as.numeric(data$Duration)
data<-data %>% mutate(Duration=abs(Duration))
data
```
```{r}
#Omitting Na values of Unnecessary rows
data_1<- data %>% 
  filter(!is.na(Date_Started)&!is.na(Date_Ended))
View(data)
```
### Inspecting and Hundling Missing Values with Mean for Numeric Columns and Mode for non numeric columns:
```{r}
colSums(is.na(data_1))
```
```{r}
# Function to calculate mode
calculate_mode <- function(x) {
  uniq_x <- unique(na.omit(x))
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

# Apply mean for numeric columns and mode for non-numeric columns
data_2 <- data_1 %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  mutate(across(where(Negate(is.numeric)), ~ ifelse(is.na(.), calculate_mode(.), .)))

```
### Checking Missing Values after Imputation:
```{r}
colSums(is.na(data_2))
```
### Checking and Hundling Outliers:
#### 1)Visualizing the Outliers with boxplot:
```{r}
ggplot(data_2,aes(y=Duration)) + geom_boxplot()
```
#### 2) Hundling Outliers with Lowerbound and Upperbounds If the distribution is naturally follows Normal Distribution: 
But for now lets just use the dataset with outliers.
```{r}
# Calculate and cap outliers using ifelse
data_3 <- data_2%>%
  mutate(across(where(is.numeric), ~ {
    iqr <- IQR(.x, na.rm = TRUE)
    lower_bound <- quantile(.x, 0.25, na.rm = TRUE) - 1.5 * iqr
    upper_bound <- quantile(.x, 0.75, na.rm = TRUE) + 1.5 * iqr
    # Use ifelse to cap values to the lower and upper bounds
    ifelse(.x < lower_bound, lower_bound, ifelse(.x > upper_bound, upper_bound, .x))
  }))
```
#### 3) Visualizing after Hundling Outliers:

```{r}
ggplot(data_3,aes(y=Duration))+geom_boxplot()
```
### Visualizing the distribution of Patient Data Using histogram and Density Plot:
Since we are not sure the type of distribution for our dataset lets skip to handle the outliers.

```{r}
# Step 1: Visualize the data
ggplot(data_2, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(alpha = .2, fill = "red") +
  labs(title = "Histogram and Density Plot of Patient data", x = "Duration", y = "Density") 
```
```{r}
# Step 1: Visualize the data
ggplot(data_2, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(color = "red",size=1) +
  labs(title = "Histogram and Density Plot of Patient data", x = "Duration", y = "Density")
```


```{r}
# Fit distributions
fit_normal <- fitdist(data_2$Duration, "norm")
fit_gamma <- fitdist(data_2$Duration, "gamma")
fit_weibull <- fitdist(data_2$Duration, "weibull")
fit_lognormal <- fitdist(data_2$Duration, "lnorm")
fit_exponential <- fitdist(data_2$Duration, "exp")
fit_cauchy <- fitdist(data_3$Duration, "cauchy")

# Compare Goodness-of-Fit statistics
gof <- gofstat(list(fit_normal, fit_gamma, fit_weibull, fit_lognormal, fit_exponential,fit_cauchy))
# Print AIC and BIC values
print(gof$aic)  # AIC values for each distribution
print(gof$bic)  # BIC values for each distribution

# Create a sequence of x values for plotting the fitted distributions
x_values <- seq(min(data_2$Duration, na.rm = TRUE), max(data_2$Duration, na.rm = TRUE), length.out = 100)
# Create histogram and density plot with fitted distributions
ggplot(data_2, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  
  # Add density curve and include it in the legend
  geom_density(aes(color = "Density"), size = 1) +
  
  # Add fitted distribution curves
  stat_function(fun = dgamma, args = list(shape = fit_gamma$estimate["shape"], rate = fit_gamma$estimate["rate"]),
                aes(color = "Gamma"), size = 1, linetype = "solid") +
  stat_function(fun = dweibull, args = list(shape = fit_weibull$estimate["shape"], scale = fit_weibull$estimate["scale"]),
                aes(color = "Weibull"), size = 1, linetype = "solid") +
  stat_function(fun = dnorm, args = list(mean = fit_normal$estimate["mean"], sd = fit_normal$estimate["sd"]),
                aes(color = "Normal"), size = 1, linetype = "solid") +
  stat_function(fun = dlnorm, args = list(meanlog = fit_lognormal$estimate["meanlog"], sdlog = fit_lognormal$estimate["sdlog"]),
                aes(color = "Lognormal"), size = 1, linetype = "solid") +
  stat_function(fun = dexp, args = list(rate = 1 / mean(data_2$Duration, na.rm = TRUE)),
                aes(color = "Exponential"), size = 1, linetype = "solid") +
  
  # Labels and theme adjustments
  labs(title = "Histogram and Density Plot of Duration with Fitted Distributions",
       x = "Duration", y = "Density") +
  theme_minimal() +
  
  # Define legend colors
  scale_color_manual("Distributions", 
                     values = c("Density" = "red", "Gamma" = "black", "Normal" = "purple", 
                                "Weibull" = "orange", "Lognormal" = "green", "Exponential" = "blue")) +
  
  theme(legend.position = "right", legend.title = element_text(size = 12))

```

### Fitting the patient data to Beta Distribution:
```{r}
# Ensure your Duration data is in the [0, 1] range
data_2$Duration_normalized <- (data_2$Duration - min(data_2$Duration, na.rm = TRUE)) / 
                              (max(data_2$Duration, na.rm = TRUE) - min(data_2$Duration, na.rm = TRUE))

# Handling zeros and ones by adding a small value
epsilon <- 1e-5
data_2$Duration_normalized <- ifelse(data_2$Duration_normalized == 0, epsilon, data_2$Duration_normalized)
data_2$Duration_normalized <- ifelse(data_2$Duration_normalized == 1, 1 - epsilon, data_2$Duration_normalized)

# Fit Beta distribution
fit_beta <- fitdist(data_2$Duration_normalized, "beta", start = list(shape1 = 1, shape2 = 1))
summary(fit_beta)

# Create a sequence of x values for plotting the fitted Beta distribution
x_values <- seq(0, 1, length.out = 100)

# Create histogram and density plot with Beta fit
ggplot(data_2, aes(x = Duration_normalized)) +
  geom_histogram(aes(y = ..density.. * 1.5), bins = 30, fill = "lightblue", color = "black") +  # Increase the height by a factor of 1.5
  geom_density(color = "red", size = 1) +  # Density of the data
  stat_function(fun = dbeta, 
                args = list(shape1 = fit_beta$estimate["shape1"], shape2 = fit_beta$estimate["shape2"]),
                color = "blue", size = 1, linetype = "solid") +  # Fitted Beta distribution
  labs(title = "Histogram and Density Plot of Patients with Fitted Beta Distribution",
       x = "Normalized Duration", y = "Density") +
  theme_minimal()
```



```{r}
# Fit T-distribution
library(MASS)
fit_t <- fitdistr(data_3$Duration, "t")

# Plotting T-distribution
ggplot(data_3, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "steelblue", color = "white") +
  geom_density(color = "red", size = 1) +
  stat_function(fun = dt, args = list(df = fit_t$estimate["df"]),
                color = "darkred", size = 1, linetype = "solid") +
  labs(title = "Histogram and Density Plot with Fitted T-Distribution",
       x = "Duration", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title

```

```{r}
# Fit Cauchy distribution
fit_cauchy <- fitdist(data_3$Duration, "cauchy")
# Plotting Cauchy distribution
ggplot(data_3, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "steelblue", color = "white") +
  geom_density(color = "red", size = 1) +
  stat_function(fun = dcauchy, args = list(location = fit_cauchy$estimate["location"], scale = fit_cauchy$estimate["scale"]),
                color = "darkred", size = 1, linetype = "solid") +
  labs(title = "Histogram and Density Plot with Fitted Cauchy Distribution",
       x = "Duration", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))  # Center the title
```


### Fitting the Pareto distribution Manually with the density plot:

```{r}
# 1. Define the Pareto PDF
dpareto_manual <- function(x, shape, location) {
  ifelse(x < location, 0, (shape * location^shape) / (x^(shape + 1)))
}

# 2. Calculate the MLE for the shape parameter
pareto_mle_shape <- function(data) {
  n <- length(data)
  shape_mle <- n / sum(log(data / min(data)))  # MLE for shape
  return(shape_mle)
}

# Calculate MLE for shape
shape_mle <- pareto_mle_shape(data_2$Duration)

# 3. Set the location parameter as the minimum value of the Duration data
location_mle <- min(data_2$Duration, na.rm = TRUE)

# Print the estimated parameters
cat("Estimated shape parameter (MLE):", shape_mle, "\n")
cat("Estimated location parameter (MLE):", location_mle, "\n")

# 4. Create a sequence of x values for plotting the fitted distribution
x_values <- seq(location_mle, max(data_2$Duration, na.rm = TRUE), length.out = 100)

# 5. Create histogram and overlay the fitted Pareto distribution
ggplot(data_2, aes(x = Duration)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(color = "red", size = 1) +
  stat_function(fun = dpareto_manual, 
                args = list(shape = shape_mle, location = location_mle),
                color = "green", size = 1, linetype = "solid") +
  labs(title = "Histogram and Fitted Pareto Distribution",
       x = "Duration", y = "Density") +
  theme_minimal()
```
```{r}
# Load necessary libraries
library(ggplot2)
library(fitdistrplus)

# Ensure your Duration data is in the [0, 1] range
data_2$Duration_normalized <- (data_2$Duration - min(data_2$Duration, na.rm = TRUE)) / 
                              (max(data_2$Duration, na.rm = TRUE) - min(data_2$Duration, na.rm = TRUE))

# Handling zeros and ones by adding a small value
epsilon <- 1e-5
data_2$Duration_normalized <- ifelse(data_2$Duration_normalized == 0, epsilon, data_2$Duration_normalized)
data_2$Duration_normalized <- ifelse(data_2$Duration_normalized == 1, 1 - epsilon, data_2$Duration_normalized)

# Fit Beta distribution
fit_beta <- fitdist(data_2$Duration_normalized, "beta", start = list(shape1 = 1, shape2 = 1))
summary(fit_beta)

# Create a sequence of x values for plotting the fitted Beta distribution
x_values <- seq(0, 1, length.out = 100)

# Create histogram and density plot with Beta fit
ggplot(data_2, aes(x = Duration_normalized)) +
  geom_histogram(aes(y = ..density.. * 1.5), bins = 30, fill = "lightblue", color = "black") +  # Increase the height by a factor of 1.5
  geom_density(color = "red", size = 1) +  # Density of the data
  
  stat_function(fun = dbeta, 
                args = list(shape1 = fit_beta$estimate["shape1"], shape2 = fit_beta$estimate["shape2"]),
                color = "blue", size = 1, linetype = "solid") +  # Fitted Beta distribution
  labs(title = "Histogram and Density Plot of Patients with Fitted Beta Distribution",
       x = "Normalized Duration", y = "Density") +
  theme_minimal()

```
```{r}
# Load necessary libraries
library(fitdistrplus)  # Ensure this library is loaded for fitdist
library(dplyr)

# Function to extract relevant statistics including log-likelihood
get_summary <- function(fit) {
  c(AIC = fit$aic, BIC = fit$bic, 
    logLik = logLik(fit),  # Extract log-likelihood
    mean = ifelse("mean" %in% names(fit$estimate), fit$estimate["mean"], NA),
    sd = ifelse("sd" %in% names(fit$estimate), fit$estimate["sd"], NA),
    shape = ifelse("shape" %in% names(fit$estimate), fit$estimate["shape"], NA),
    scale = ifelse("scale" %in% names(fit$estimate), fit$estimate["scale"], NA),
    meanlog = ifelse("meanlog" %in% names(fit$estimate), fit$estimate["meanlog"], NA),
    sdlog = ifelse("sdlog" %in% names(fit$estimate), fit$estimate["sdlog"], NA),
    shape1 = ifelse("shape1" %in% names(fit$estimate), fit$estimate["shape1"], NA),
    shape2 = ifelse("shape2" %in% names(fit$estimate), fit$estimate["shape2"], NA)
  )
}

# Collect summaries for all fitted distributions
summary_stats <- data.frame(
  Distribution = c("Normal", "Gamma", "Weibull", "Lognormal", "Exponential", "Beta", "Pareto"),
  rbind(
    get_summary(fit_normal),
    get_summary(fit_gamma),
    get_summary(fit_weibull),
    get_summary(fit_lognormal),
    get_summary(fit_exponential),
    get_summary(fit_beta),  # Assuming fit_beta is fitted
    c(AIC = NA, BIC = NA, 
      logLik = logLik_pareto,  # Replace with the calculated log-likelihood for Pareto
      mean = location_mle,      # Location for Pareto
      sd = NA,                  # No standard deviation for Pareto
      shape = shape_mle,        # Shape for Pareto
      scale = NA,               # No scale for Pareto
      meanlog = NA,
      sdlog = NA,
      shape1 = NA,
      shape2 = NA)
  )
)

# Print the summary statistics
print(summary_stats)

```




